---
title: "Merck-Datathon"
author: "Zijun Liu"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: haddock
    theme: default
    code_folding: hide
    df_print: paged
    toc: yes
    toc_depth: '3'
    toc_float:
      collapsed: yes
      smooth_scroll: yes
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(paste0("~/Dropbox/Merck-Datathon/05_output/Merck-Datathon-EDA-and-Analysis-", Sys.Date(), ".html"))) })
---

```{r}
library(tidyverse)
library(table1)
library(arrow) # to read Apache parquet data
library(here)
library(broom)
library(survival)
library(survminer)
library(readxl)
library(grid)
library(forestploter)
here()
```

State level
```{r}
# read state data: each row is a state-year
df_state <- read_parquet(here("03_clean_data", "State_FinalizedData.parquet"))

# platform entry of each state: find thee min for each state in the two files
df_entry_2016 <- read_csv(here("02_raw_data", "Section1115_SUD_States_By_Year_2016_inception.csv"))
df_entry_2017 <- read_csv(here("02_raw_data", "Section1115_SUD_States_By_Year_2017_inception.csv"))

# fips-state link
df_fips <- read_csv(here("02_raw_data", "FIPS_Mapping.csv"))

# data for fig1. forest plot of subgroup coefficients
df_fig1 <- read_excel(here("03_clean_data", "Consolidated Tables.xlsx"),
                      sheet = "fig1")
```


```{r}
control_cols = c('log_population','male_prop', 'age_15_24_prop', 'age_25_44_prop', 'age_45_64_prop',
                'age_65_prop', 'white_prop', 'black_prop', 'asian_prop',
                'hispanic_prop', 'median_income', 'average_income', 'unemploy_prop',
                'no_insurance_prop', 'high_school_prop', 'college_prop',
                'marriage_prop', 'population', 'state_abbr', 'total_drug_tot_arrests',
                'disorder_cond_tot_arrests', 'sale_drug_total_tot_arrests',
                'poverty_all_ages', 'median_household_income')

# only use some cols from above in diff-in-diff model
control_cols_DID = c('male_prop','white_prop', 'age_15_24_prop','median_income','unemploy_prop', 'log_population', 'total_drug_tot_arrests')


treatment_cols = c('pr_abuse_med_pct',
       'pr_abuse_med_pay_pct', 'EMPLOY_Improved', 'LIVARAG_Improved',
       'ARRESTS_Improved', 'DAYWAIT_Extreme', 'Treatment_completed',
       'Treatment_death', 'FREQ1_Improved', 'FREQ_Improved',
       'EMPLOY_Improved_withWage', 'LIVARAG_Improved_withWage',
       'ARRESTS_Improved_withWage', 'DAYWAIT_Extreme_withWage',
       'Treatment_completed_withWage', 'Treatment_death_withWage',
       'FREQ1_Improved_withWage', 'FREQ_Improved_withWage',
       'EMPLOY_Improved_woman', 'LIVARAG_Improved_woman',
       'ARRESTS_Improved_woman', 'DAYWAIT_Extreme_woman',
       'Treatment_completed_woman', 'Treatment_death_woman',
       'FREQ1_Improved_woman', 'FREQ_Improved_woman',
       'EMPLOY_Improved_african', 'LIVARAG_Improved_african',
       'ARRESTS_Improved_african', 'DAYWAIT_Extreme_african',
       'Treatment_completed_african', 'Treatment_death_african',
       'FREQ1_Improved_african', 'FREQ_Improved_african',
       'EMPLOY_Improved_preg', 'LIVARAG_Improved_preg',
       'ARRESTS_Improved_preg', 'DAYWAIT_Extreme_preg',
       'Treatment_completed_preg', 'Treatment_death_preg',
       'FREQ1_Improved_preg', 'FREQ_Improved_preg', 'EMPLOY_Improved_co',
       'LIVARAG_Improved_co', 'ARRESTS_Improved_co', 'DAYWAIT_Extreme_co',
       'Treatment_completed_co', 'Treatment_death_co', 'FREQ1_Improved_co',
       'FREQ_Improved_co', 'LOS_STAY_withWage', 'LOS_STAY', 'LOS_STAY_preg')

facility_cols = c('REVCHK5_sum', 'TREATMT_sum', 'REVCHK5_gov_sum',
       'TREATMT_gov_sum', 'ASSESSMENT_withTREATMT_mean',
       'CTYPE7_withTREATMT_sum', 'REVCHK5_cplx_sum', 'REVCHK5_withTREATMT_sum',
       'REVCHK5_sum_per_pop', 'TREATMT_sum_per_pop', 'REVCHK5_gov_sum_per_pop',
       'TREATMT_gov_sum_per_pop', 'CTYPE7_withTREATMT_sum_per_pop',
       'REVCHK5_cplx_sum_per_pop', 'REVCHK5_withTREATMT_sum_per_pop')

# CTYPE7: Facility offers non-hospital residential treatment services
target_col = 'CTYPE7_withTREATMT_sum'
event_col = 'Event2016'
time_col = 'year'
id_col = 'state_abbr'
```

Questions: 'total_drug_tot_arrests' , 'Event2016', 'DAYWAIT_Extreme' meaning?
what is daywait variable in state level?

# 1. Table 1
```{r}
# diff-in-diff model
# reg_did_df = reg_df[ [time_col,id_col,target_col,event_col]+used_cols ].copy().dropna()
table1_col = c('log_population', 'male_prop', 'white_prop', 'age_15_24_prop', 'age_65_prop', 'median_income','unemploy_prop', 'total_drug_tot_arrests', 'poverty_all_ages')

# df_state[c('year', 'Event2016', 'TREATMT_sum', 'TREATMT_sum_per_pop', 'DAYWAIT_Extreme', 'CTYPE7_withTREATMT_sum')]

label(df_state$log_population) <- "log(Population)"
label(df_state$male_prop) <- "Male Proportion"
label(df_state$white_prop) <- "White Proportion"
label(df_state$age_15_24_prop) <- "Age 15-24 Proportion"
label(df_state$age_65_prop) <- "Age 65+ Proportion"
label(df_state$median_income) <- "Income"
label(df_state$unemploy_prop) <- "Unemployment Proportion"
label(df_state$total_drug_tot_arrests) <- "Arrest due to Drug Proportion"
label(df_state$poverty_all_ages) <- "Poverty Proportion"

units(df_state$median_income) <- "Dollars"

table1(~ log_population + male_prop + white_prop + age_15_24_prop + 
       age_65_prop + median_income + unemploy_prop + total_drug_tot_arrests +
         poverty_all_ages| year, 
       data=df_state,
       caption = "Descriptive statistics by year")
```
# 2. Cox regression as a robustness check
Predict the platform entry as a function of drug abuse and other independent variables



**Question: if state not found in both files, state had not entered as of 2020?**
```{r cox data cleaning}
# colnames(df_entry_2016)
# rename coumns for merging
df_entry_2016 <- df_entry_2016 %>% 
  rename(state = IMD_Payment_Exclusion_SUD_Treatment,
         year = Year) %>% 
  select(-Link)

df_entry_2017 <- df_entry_2017 %>% 
  rename(state = IMD_Payment_Exclusion_SUD_Treatment,
         year = Year) %>% 
  select(-Link)

# find the entry time for each state
# if no entry time, state had not entered as of 2020
df_entry_year <- df_entry_2016 %>% 
  full_join(df_entry_2017, by = "state") %>% 
  # find the smaller year in two df
  mutate(year = ifelse(year.x<=year.y, year.x, year.y)) %>% 
  # find the smallest year for each state
  group_by(state) %>% 
  summarise(entry_year = min(year))

# add fips code for the state
df_entry_year <- df_fips %>% 
  left_join(df_entry_year, by = c("STATE"="state")) %>% 
  rename(state=STATE)
```


```{r}
# build data frame for survival analysis
df_cox1 <- df_entry_year %>% 
         # entry = o: censored, entry=1: censored/died
  mutate(entry = ifelse(is.na(entry_year), 0, 1),
         # time: survival time in years
         time = ifelse(is.na(entry_year), 2020-2015, entry_year-2015))

# extract X variables for Cox regression from df_state
# and add it to df_cox1
df_cox2 <- df_state %>% 
  filter(year == 2015) %>% 
  select(state_fips, FREQ1_Improved, FREQ1_Improved_co, log_population,male_prop, white_prop, 
         age_15_24_prop, age_65_prop, median_income, unemploy_prop,
         total_drug_tot_arrests, poverty_all_ages) %>% 
  left_join(df_cox1, by = "state_fips") %>% 
  select(state_fips, state, entry_year, entry, time, everything())
```


```{r km plot, include=FALSE}
km_fit <- survfit(Surv(time, entry) ~ 1,
                  data=df_cox2)
# life table
summary(km_fit)

# km plot
ggsurvplot(km_fit, data=df_cox2,
           conf.int=TRUE, 
           surv.median.line = "hv",
           #risk.table="abs_pct", 
           xlab = "Years before state entry of Section 1115",
           title="Kaplan-Meier Curve for Section 1115 - state level", 
           risk.table.height=.3, fontsize=3)
```

## 2.1 model FREQ1_Improved

Cox regression - Breslow to handle ties
```{r cox}
# cox model with all variables
m.cox.breslow <- coxph(Surv(time, entry) ~ FREQ1_Improved + log_population + male_prop + 
                 white_prop + age_15_24_prop + age_65_prop + median_income +
                 unemploy_prop + total_drug_tot_arrests + poverty_all_ages,
               data = df_cox2, 
               ties = "breslow")

# summary(m.cox.breslow)
tidy(m.cox.breslow)
```

Cox regression - Efron to handle ties
```{r}
# cox model with all variables
m.cox.efron <- coxph(Surv(time, entry) ~ FREQ1_Improved + log_population + male_prop + 
                 white_prop + age_15_24_prop + age_65_prop + median_income +
                 unemploy_prop + total_drug_tot_arrests + poverty_all_ages,
               data = df_cox2, 
               ties = "efron")

# summary(m.cox.efron)
tidy(m.cox.efron)
```

Cox regression - Exact partial likelihood to handle ties
```{r}
# cox model with all variables
m.cox.exact <- coxph(Surv(time, entry) ~ FREQ1_Improved + log_population + male_prop + 
                 white_prop + age_15_24_prop + age_65_prop + median_income +
                 unemploy_prop + total_drug_tot_arrests + poverty_all_ages,
               data = df_cox2, 
               ties = "exact")

# summary(m.cox.exact)
tidy(m.cox.exact)
```


## 2.2 model FREQ1_Improved_co

Cox regression - Breslow to handle ties
```{r}
# cox model with all variables
m.cox.breslow <- coxph(Surv(time, entry) ~ FREQ1_Improved_co + log_population + male_prop + 
                 white_prop + age_15_24_prop + age_65_prop + median_income +
                 unemploy_prop + total_drug_tot_arrests + poverty_all_ages,
               data = df_cox2, 
               ties = "breslow")

# summary(m.cox.breslow)
tidy(m.cox.breslow)
```

Cox regression - Efron to handle ties
```{r}
# cox model with all variables
m.cox.efron <- coxph(Surv(time, entry) ~ FREQ1_Improved_co + log_population + male_prop + 
                 white_prop + age_15_24_prop + age_65_prop + median_income +
                 unemploy_prop + total_drug_tot_arrests + poverty_all_ages,
               data = df_cox2, 
               ties = "efron")

# summary(m.cox.efron)
tidy(m.cox.efron)
```

Cox regression - Exact partial likelihood to handle ties
```{r}
# cox model with all variables
m.cox.exact <- coxph(Surv(time, entry) ~ FREQ1_Improved_co + log_population + male_prop + 
                 white_prop + age_15_24_prop + age_65_prop + median_income +
                 unemploy_prop + total_drug_tot_arrests + poverty_all_ages,
               data = df_cox2, 
               ties = "exact")

# summary(m.cox.exact)
tidy(m.cox.exact)
```

# 3. fig1

```{r fig.width=10, fig.height=10, eval=FALSE}

# Read provided sample example data
dt <- read.csv(system.file("extdata", "example_data.csv", package = "forestploter"))

# Keep needed columns
dt <- dt[,1:6]

# indent the subgroup if there is a number in the placebo column
dt$Subgroup <- ifelse(is.na(dt$Placebo), 
                      dt$Subgroup,
                      paste0("   ", dt$Subgroup))

# NA to blank or NA will be transformed to carachter.
dt$Treatment <- ifelse(is.na(dt$Treatment), "", dt$Treatment)
dt$Placebo <- ifelse(is.na(dt$Placebo), "", dt$Placebo)
dt$se <- (log(dt$hi) - log(dt$est))/1.96

# Add blank column for the forest plot to display CI.
# Adjust the column width with space, increase number of space below 
# to have a larger area to draw the CI. 
dt$` ` <- paste(rep(" ", 20), collapse = " ")

# Create confidence interval column to display
dt$`HR (95% CI)` <- ifelse(is.na(dt$se), "",
                             sprintf("%.2f (%.2f to %.2f)",
                                     dt$est, dt$low, dt$hi))

p <- forest(dt[,c(1:3, 8:9)],
            est = dt$est,
            lower = dt$low, 
            upper = dt$hi,
            sizes = dt$se,
            ci_column = 4,
            ref_line = 1,
            arrow_lab = c("Placebo Better", "Treatment Better"),
            xlim = c(0, 4),
            ticks_at = c(0.5, 1, 2, 3),
            footnote = "This is the demo data. Please feel free to change\nanything you want.")

# Print plot
p
```

```{r}
# prepare data for forest plot
df_fig1_tidy <- df_fig1 %>% 
  mutate(low_ci = coef-1.96*se,
         high_ci = coef+1.96*se,
         Subgroup = str_replace(subgroup_label, "All Population: ", ""),
         Subgroup = str_replace(Subgroup, "Age group \\d+: ", ""),
         ` ` = paste(rep(" ", 20), collapse = " "))
df_fig1_tidy[df_fig1_tidy$subgroup=="medicaid", ]$Subgroup <- "Medicaid"

# forest plot
forest(df_fig1_tidy[,c(8,9)],
            est = df_fig1_tidy$coef,
            lower = df_fig1_tidy$low_ci, 
            upper = df_fig1_tidy$high_ci,
            sizes = 0.5,
            ci_column = 2,
            ref_line = 0,
            #arrow_lab = c("Placebo Better", "Treatment Better"),
            xlim = c(-0.05, 0.12),
            ticks_at = c(-0.15, -0.1, -0.05, 0, 0.05, 0.1, 0.15))

```
